---
title: "CEMA Internship Task"
author: "Humphrey Kinoti"
date: "`r Sys.Date()`"
output: pdf_document
editor_options: 
  markdown: 
    wrap: 72
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = F, warning = F)
```

***A copy of this project is in:*** [link to my github
repo](https://github.com/HumphreyKn/internship_task)

## Instructions

You have been provided with a dataset which contains monthly data for
children \<5 years, disaggregated at a county level for the period
January 2021 to June 2023.

### Dataset description

The dataset contains the following variables:

-   `period`: months from January 2021 to June 2023

-   `county`: the 47 counties in Kenya

-   `Total Dewormed`: Total number of children dewormed

-   `Acute Malnutrition`: Number of children \<5 years with acute
    malnutrition

-   `stunted 0-6 months`, `stunted 6-23 months`, `stunted 24-59 months`:
    Number of children stunted

-   `diarrhea cases`: Number of children \<5 years with diarrhea

-   `underweight 0-6 months`, `underweight 6-23 months`,
    `underweight 24-59 months`: Number of children who are underweight

### Objectives

Your task is to: -

-   Conduct exploratory data analysis

-   State an appropriate research question you would want to answer from
    the data

-   Carry out appropriate data analysis to address the question you have
    stated above

## Explanatory Data Analysis

### Import Data
As the data is in CSV format, we will import data using `read_csv()` function in the `readr` package.

```{r}
library(tidyverse)
cema_data <- read_csv("data/cema_internship_task_2023.csv")
```

After the importation, Let's have a look of how the data is structured to have a better understanding of the data. In this case, I will use the `skim` function from `skimr` package.

```{r}
library(skimr)
skim(cema_data) %>% 
  dplyr::select(skim_type, skim_variable, n_missing, 
                character.min, character.max, character.empty,
                character.n_unique,
                numeric.mean, numeric.sd, numeric.p0, numeric.p25, numeric.p50, numeric.p75, numeric.p100)
```

